Calibrating MEM seeding heuristics

### 2018-04-24 MEM.tex
Text is a mess but formula are mostly correct except typos.

\begin{definition}
$\xi_{i,m}$ is the probability that at least one of $m$ alternative
threads does not fail in $i$ nucleotides, \textit{i.e.}
\begin{equation}
  \xi_{i,m} = 1-(1-(1-\mu)^i)^m.
\end{equation}
\end{definition}

@@

If the next terminator is $\Downarrow$, there is a limit on the
number of $\square$ symbols. Indeed, if there are $\gamma$ or more
nucleotides between the two $\Downarrow$ symbols, they constitute a shared
MEM seed. As mentioned in section~\ref{sec:trans_from_up}, the weighted
generating function of the $\Downarrow$ symbol is $pz(1-\mu/3)^N$. By the
same rationale as above, the weighted generating function of the
transition from $\Downarrow$ to $\Downarrow$ is
\begin{equation}
A_{\gamma}(z) = \omega z
  \sum_{i=0}^{\gamma-1} \xi_{i,N} \cdot (qz)^i.
\end{equation}

If the next terminator is $\downarrow$, there is no limit on the number of
$\square$ symbols that can be inserted. Indeed, the main thread will not
be part of the MEM in this case. As mentioned in
section~\ref{sec:trans_from_up}, the weighted generating function of the
$\downarrow$ symbol is $pz(1-(1-\mu/3)^N)$. The probability that it first
occurs at position $n+1$ after the $\Downarrow$ symbol is the probability
that the first $n$ nucleotides are correct -- each with probability $q$
-- and that at least one alternative thread survives -- probability
$\xi_{n,N}$. The weighted generating function of the transition from
$\Downarrow$ to $\downarrow$ is thus
\begin{equation}
\tilde{A}_\infty(z) = \tilde{\omega} z
  \sum_{i=0}^\infty \xi_{i,N} \cdot (qz)^i.
\end{equation}

@@

\begin{eqnarray}
\alpha_i &=& 1-(1-\mu)^i\mu/3 \\
\gamma_i &=& 1-(1-\mu)^i \\
\delta_i &=& 1-(1-\mu+\mu^2/3)(1-\mu)^i.
\end{eqnarray}

We have assumed $m>0$, but the sum above can start at $m=0$ because
$\xi_{i,0}=0$. So the weighted generating function of the transition from
$\downarrow$ to $(\sim)_i$ is
\begin{equation}
v_i(z) = \frac{\alpha_i^N-\alpha_{i-1}^N-
  \gamma_i^N+\delta_{i-1}^N}{1-(1-\mu/3)^N} \cdot (qz)^i.
\end{equation}

@@

In order to compute the weighted generating functions of the other
transitions, first observe that
\begin{equation}
V_j+W_j = \frac{\alpha_j^N - \alpha_{j-1}^N}{1-(1-\mu/3)^N}.
\end{equation}
So the probability that an error-free stretch of length $i \geq 1$
following a $\downarrow$ symbol contains neither a $\sim$ symbol nor a
$\uparrow$ symbol is
\begin{equation}
\sum_{j=1}^i V_j+W_j = \frac{\alpha_i^N -
\alpha_0^N}{1-(1-\mu/3)^N}.
\end{equation}

The probability that the next terminator after a $\downarrow$ symbol is a
$\Downarrow$ symbol at position $j$ is the probability of an error-free
stretch of size $j-1$ that contains neither the $\sim$ symbol nor the
$\uparrow$ symbol -- probability as above multiplied by $q^j$ -- followed
by a $\Downarrow$ symbol -- probability $pz(1-\mu/3)^N$. Thus, the
weighted generating function of the transition from $\downarrow$ to
$\Downarrow$ is
\begin{equation}
B_\infty(z) = \omega z \cdot \Big(1 + \sum_{i=1}^\infty
\frac{\alpha_i^N - \alpha_0^N}{1-(1-\mu/3)^N} (qz)^i \Big).
\end{equation}

By the same rationale, we see that the weighted generating function of the
transition from $\downarrow$ to $\downarrow$ is
\begin{equation}
\tilde{B}_\infty(z) = \tilde{\omega} z \cdot \Big(1 + \sum_{i=1}^\infty
\frac{\alpha_i^N - \alpha_0^N}{1-(1-\mu/3)^N} (qz)^i \Big).
\end{equation}

@@

The probability that there are $m>1$ potentially masking threads upon
entering state $(\sim)_j$ is the probability that there were $n \geq m$
potentially masking threads at the $\downarrow$ symbol -- probability
$Q_n$ --, that exactly $m$ have survived until position $j$ -- probability
${n \choose m} (1-\mu)^{jm} (1-(1-\mu)^j)^{n-m}$ -- and that the last of
the $N-n$ strictly masking threads fails at position $j$ -- probability
$\xi_{j-1,N-n}-\xi_{j,N-n}$. Summing over $n = m, \ldots, N-1$, the
probability is seen to be
\begin{equation}
\label{eq:miscprob}
\sum_{n=m}^{N-1} Q_n \cdot {n \choose m} (1-\mu)^{jm} (1-(1-\mu)^j)^{n-m}
\cdot (\xi_{j-1,N-n}-\xi_{j,N-n}).
\end{equation}

@@

The probability that the last of the $m$ potentially masking threads fails
at position $i$ after entering state $(\sim)_j$ is
$\xi_{i-1,m}-\xi_{i,m}$. Dividing by $V_j$ and summing over $m = 1,
\ldots, N-1$, we find that the weighted generating function of the
transition from $(\sim)_j$ to $(\uparrow)_{j+i}$ is
\begin{eqnarray}
y_{j,j+i}(z) &=& \sum_{m=1}^{N-1} \frac{P_{m,j}}{V_j}
\big( \xi_{i-1,m}-\xi_{i,m} \big) (qz)^i \notag \\
&=& \frac{\beta_{j,i+j}^N - \beta_{j,i+j-1}^N
-\beta_{j-1,i+j}^N + \beta_{j-1,i+j-1}^N}
  {\alpha_j^N-\alpha_{j-1}^N-\gamma_j^N+\delta_{j-1}^N} (qz)^i,
\end{eqnarray}
where $\alpha_j$, $\delta_j$, $\gamma_j$ are as above, and where
\begin{equation}
\beta_{j,i} = 1 - (1-\mu)^j\mu/3 + (1-\mu)^i(1-\mu/3).
\end{equation}

@@

For the $\downarrow$ symbol, the weighted generating function is simply
\begin{equation}
T_{\downarrow}(z) = \sum_{i=0}^\infty (qz)^i = \frac{1}{1-qz}.
\end{equation}

@@

In summary,
\begin{equation*}
T(z) = 
\begin{blockarray}{cc}
\begin{block}{c[c]}
\dn & T_\downarrow(z) \\
\DN & T_\Downarrow(z) \\
\eq{1} & T_{(\sim)_1}(z) \\
\vdots & \vdots \\
\eq{\gamma-1} & T_{(\sim)_{\gamma-1}}(z) \\
\up{1} & T_{(\uparrow)_1}(z) \\
\vdots & \vdots \\
\up{\gamma-1} & \; T_{(\uparrow)_{\gamma-1}}(z) \; \\
\end{block}
\end{blockarray}.
\end{equation*}

@@

\begin{equation*}
\begin{blockarray}{ccccccccc}
   & \DN & \dn &\eq{1} & \ldots & \eq{\gamma-1} &
    \up{1} & \ldots & \up{\gamma-1} \\
\begin{block}{c[cccccccc]}
\DN & A_\gamma(z) & \tilde{A}_\infty(z) & 0 & \ldots & 0 & u_1(z)
    & \ldots & u_{\gamma-1}(z) \\
\dn & B_\infty(z) & \tilde{B}_\infty(z) & v_1(z) & \ldots &
    v_{\gamma-1}(z) & w_1(z) & \ldots & w_{\gamma-1}(z) \\
\eq{1} & C_{\gamma-1}(z) & \tilde{C}_{\gamma-1}(z) & 0 & \ldots & 0 & 0 &
    \ldots & y_{1,\gamma-1}(z) \\
\eq{2} & C_{\gamma-2}(z) & \tilde{C}_{\gamma-2}(z) & 0 & \ldots & 0 & 0 &
    \ldots & y_{2,\gamma-1}(z) \\
\vdots & \vdots & \vdots & \vdots & \ddots & \vdots & \vdots &
    \ddots & \vdots \\
\eq{\gamma-1} & C_1(z) & \tilde{C}_1(z) & 0 & 
    \ldots & 0 & 0 & \ldots & 0 \\
\up{1} & D_{\gamma-1}(z) & \tilde{D}_{\gamma-1}(z) & 0 & \ldots & 0 & 0 &
    \ldots & 0 \\
\up{2} & D_{\gamma-2}(z) & \tilde{D}_{\gamma-2}(z) & 0 & \ldots & 0 & 0 &
    \ldots & 0  \\
\vdots & \vdots & \vdots & \vdots & \ddots & \vdots & \vdots &
    \ddots & \vdots \\
\up{\gamma-1} & D_1(z) & \tilde{D}_1(z) & 0 & \ldots & 0 & 0 &
    \ldots & 0 \\
\end{block}
\end{blockarray}
\end{equation*}

>>>>

\begin{definition}
$\xi_{i,m}$ is the probability that at least one of $m$ alternative
threads does not fail in $i$ error-free nucleotides, \textit{i.e.}
\begin{equation}
  \xi_{i,m} = 1-(1-(1-\mu)^i)^m.
\end{equation}
\end{definition}

@@

If the next terminator is $\Downarrow$, there is a limit on the
number of $\square$ symbols. Indeed, if there are $\gamma$ or more
nucleotides between the two $\Downarrow$ symbols, they constitute a shared
MEM seed. As mentioned in section~\ref{sec:trans_from_up}, the weighted
generating function of the $\Downarrow$ symbol is $pz(1-\mu/3)^N$. By the
same rationale as above, the weighted generating function of the
transition from $\Downarrow$ to $\Downarrow$ is
\begin{equation}
A_\gamma(z) = \omega z
  \sum_{i=0}^{\gamma-1} \xi_{i,N} \cdot (qz)^i.
\end{equation}

If the next terminator is $\downarrow$ we have to dinstinguish two cases.
If there are $\gamma-1$ or fewer error-free nucleotides, we must only make
sure that there is no $\uparrow$ symbol among them. The weighted
generating function is
\begin{equation}
\tilde{\omega} z \sum_{i=0}^{\gamma-1} \xi_{i,N} \cdot (qz)^i.
\end{equation}

There can also be more than $\gamma-1$ error-free nucleotides, but in this
case, we must make sure that at least one potentially masking thread masks
the main thread.

\begin{definition}
$\alpha_i$ is the probability that the first thread has at least one
failure in $i$ error-free nucleotides, followed by a sequencing error,
\textit{i.e.}
\begin{equation}
\alpha_i = 1 - (1-\mu)^i\mu/3
\end{equation}
\end{definition}

The probability that at least one duplicate thread masks the main thread
if the first error occurs at position $i$ is thus $1 - \alpha_{i-1}^N$.
The weighted generatin function is
\begin{equation}
pz\sum_{i=\gamma}^\infty (1 - \alpha_i^N) \cdot (qz)^i.
\end{equation}

Finally, the weighted generating function of the transition from
$\Downarrow$ to $\downarrow$ is
\begin{equation}
\tilde{A}_\infty(z) =
\tilde{\omega} z \sum_{i=0}^{\gamma-1} \xi_{i,N} \cdot (qz)^i +
pz\sum_{i=\gamma}^\infty (1 - \alpha_i^N) \cdot (qz)^i.
\end{equation}

@@

\begin{eqnarray}
\gamma_i &=& 1-(1-\mu)^i \\
\delta_i &=& 1-(1-\mu+\mu^2/3)(1-\mu)^i.
\end{eqnarray}

We have assumed $m>0$, but the sum above can start at $m=0$ because
$\xi_{i,0}=0$. The weighted generating function of the transition from
$\downarrow$ to $(\sim)_i$ is
\begin{equation}
v_i(z) = \frac{\alpha_i^N-\alpha_{i-1}^N-
  \gamma_i^N+\delta_{i-1}^N}{1-(1-\mu/3)^N} \cdot (qz)^i.
\end{equation}

\begin{remark}
When $N = 1$, $V_i = 0$ and none of the states $(\sim)_i$ is accessible.
The symbol $\sim$ can exist only if there is more than one duplicate.
\end{remark}

@@

%In order to compute the weighted generating functions of the other
%transitions, first observe that
%\begin{equation}
%V_j+W_j = \frac{\alpha_j^N - \alpha_{j-1}^N}{1-(1-\mu/3)^N}.
%\end{equation}
%So the probability that an error-free stretch of length $i \geq 1$
%following a $\downarrow$ symbol contains neither a $\sim$ symbol nor a
%$\uparrow$ symbol is
%\begin{equation}
%1 - \sum_{j=1}^i V_j+W_j = 1 - \frac{\alpha_i^N -
%\alpha_0^N}{1-(1-\mu/3)^N}.
%\end{equation}
%
%The probability that the next terminator after a $\downarrow$ symbol is a
%$\Downarrow$ symbol at position $j$ is the probability of an error-free
%stretch of size $j-1$ that contains neither the $\sim$ symbol nor the
%$\uparrow$ symbol -- probability as above multiplied by $q^j$ -- followed
%by a $\Downarrow$ symbol -- probability $pz(1-\mu/3)^N$. Thus, the
%weighted generating function of the transition from $\downarrow$ to
%$\Downarrow$ is
%\begin{equation}
%B_\infty(z) = \omega z \sum_{i=0}^\infty
%\Big(1 - \frac{\alpha_i^N - \alpha_0^N}{1-(1-\mu/3)^N} \Big) (qz)^i.
%\end{equation}

The probability that there is neither $\sim$ nor $\uparrow$ symbol is just
the probability that at least one stricty masking thread is still alive
\begin{equation}
\sum_{m=0}^{N-1}Q_m \cdot \xi_{i,N-m} = \frac{1-\alpha_i^N}{1-(1-\mu/3)^N}
\end{equation}

So the weighted generating function of the transition from $\downarrow$ to
$\Downarrow$ is
\begin{equation}
B_\infty(z) = \omega z\sum_{i=0}^\infty
\frac{1-\alpha_i^N}{1-(1-\mu/3)^N} (qz)^i.
\end{equation}

By the same rationale, we see that the weighted generating function of the
transition from $\downarrow$ to $\downarrow$ is
\begin{equation}
\tilde{B}_\infty(z) = \tilde{\omega} z\sum_{i=0}^\infty
\frac{1-\alpha_i^N}{1-(1-\mu/3)^N} (qz)^i.
\end{equation}

@@

The probability that there are $m>1$ potentially masking threads upon
entering state $(\sim)_j$ is the probability that there were $n \geq m$
potentially masking threads at the $\downarrow$ symbol -- probability
$Q_n$ --, that exactly $m$ have survived until position $j$ -- probability
${n \choose m} (1-\mu)^{jm} (1-(1-\mu)^j)^{n-m}$ -- and that the last of
the $N-n$ strictly masking threads fails at position $j$ -- probability
$\xi_{j-1,N-n}-\xi_{j,N-n}$. Summing over $n = m, \ldots, N-1$, the
probability is seen to be
\begin{equation}
\label{eq:miscprob}
P_{m,j} = \sum_{n=m}^{N-1} Q_n \cdot {n \choose m} (1-\mu)^{jm}
(1-(1-\mu)^j)^{n-m} \cdot (\xi_{j-1,N-n}-\xi_{j,N-n}).
\end{equation}

@@

The probability that the last of the $m$ potentially masking threads fails
at position $i$ after entering state $(\sim)_j$ is
$\xi_{i-1,m}-\xi_{i,m}$. Dividing by $V_j$ and summing over $m = 1,
\ldots, N-1$, we find that the weighted generating function of the
transition from $(\sim)_j$ to $(\uparrow)_{j+i}$ is
\begin{eqnarray}
y_{j,i}(z) &=& \sum_{m=1}^{N-1} \frac{P_{m,j}}{V_j}
\big( \xi_{i-1,m}-\xi_{i,m} \big) (qz)^i \notag \\
&=& \frac{\beta_{j,i+j}^N - \beta_{j,i+j-1}^N
-\beta_{j-1,i+j}^N + \beta_{j-1,i+j-1}^N}
  {\alpha_j^N-\alpha_{j-1}^N-\gamma_j^N+\delta_{j-1}^N} (qz)^i,
\end{eqnarray}
where $\alpha_j$, $\delta_j$, $\gamma_j$ are as above, and where
\begin{equation}
\beta_{j,i} = 1 - (1-\mu)^j\mu/3 - (1-\mu)^i(1-\mu/3).
\end{equation}

@@

For the $\downarrow$ symbol, the weighted generating function is
\begin{equation}
T_{\downarrow}(z) = \sum_{i=0}^\infty
\frac{1-\alpha_i^N}{1-(1-\mu/3)^N} (qz)^i.
\end{equation}

@@

In summary,
\begin{equation*}
T(z) = 
\begin{blockarray}{cc}
\begin{block}{c[c]}
\DN & T_\Downarrow(z) \\
\dn & T_\downarrow(z) \\
\eq{1} & T_{(\sim)_1}(z) \\
\vdots & \vdots \\
\eq{\gamma-1} & T_{(\sim)_{\gamma-1}}(z) \\
\up{1} & T_{(\uparrow)_1}(z) \\
\vdots & \vdots \\
\up{\gamma-1} & \; T_{(\uparrow)_{\gamma-1}}(z) \; \\
\end{block}
\end{blockarray}.
\end{equation*}

@@

\begin{equation*}
\begin{blockarray}{ccccccccc}
   & \DN & \dn &\eq{1} & \ldots & \eq{\gamma-1} &
    \up{1} & \ldots & \up{\gamma-1} \\
\begin{block}{c[cccccccc]}
\DN & A_\gamma(z) & \tilde{A}_\infty(z) & 0 & \ldots & 0 & u_1(z)
    & \ldots & u_{\gamma-1}(z) \\
\dn & B_\infty(z) & \tilde{B}_\infty(z) & v_1(z) & \ldots &
    v_{\gamma-1}(z) & w_1(z) & \ldots & w_{\gamma-1}(z) \\
\eq{1} & C_{\gamma-1}(z) & \tilde{C}_{\gamma-1}(z) & 0 & \ldots & 0 & 0 &
    \ldots & y_{1,\gamma-2}(z) \\
\eq{2} & C_{\gamma-2}(z) & \tilde{C}_{\gamma-2}(z) & 0 & \ldots & 0 & 0 &
    \ldots & y_{2,\gamma-3}(z) \\
\vdots & \vdots & \vdots & \vdots & \ddots & \vdots & \vdots &
    \ddots & \vdots \\
\eq{\gamma-1} & C_1(z) & \tilde{C}_1(z) & 0 & 
    \ldots & 0 & 0 & \ldots & 0 \\
\up{1} & D_{\gamma-1}(z) & \tilde{D}_{\gamma-1}(z) & 0 & \ldots & 0 & 0 &
    \ldots & 0 \\
\up{2} & D_{\gamma-2}(z) & \tilde{D}_{\gamma-2}(z) & 0 & \ldots & 0 & 0 &
    \ldots & 0  \\
\vdots & \vdots & \vdots & \vdots & \ddots & \vdots & \vdots &
    \ddots & \vdots \\
\up{\gamma-1} & D_1(z) & \tilde{D}_1(z) & 0 & \ldots & 0 & 0 &
    \ldots & 0 \\
\end{block}
\end{blockarray}
\end{equation*}
